JD notes on effects

### 2022-08-10 manuscript/paper.tex
merge with Talks progress?

Plots of predicted values of an outcome against predictors (often called effect plots or prediction plots, generalized here as “outcome plots”) are often a useful way to summarize the results of a statistical model. These can be used to illustrate model uncertainty or to give a more explicit quantitative sense of how the outcome is expected to change. In generalized models with a non-linear link function, or models with a spline or polynomial response to an input variable, they can also aid in understanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.

To make an outcome plot, we use a \emph{focal} predictor on the x-axis and values of the outomcome variable on the y-axis. The resulting plot will depend on choices we make about other (non-focal) predictors. In an ordinary linear model, the non-focal choices may have a simple additive effect on the central estimate. However, when the focal predictor has interactions or non-linear response functions (e.g., spline or polynomial), non-focal choices can also affect the slope of the estimate. Additional challenges arise when dealing with “mixed” models, which incorporate random effects.

@@

The other distinction relates not to the method of calculating CIs, but to the model chosen for the plot. 
In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model that only contains terms related to that predictor, thus addressing the question: if we know (only) the focal parameter, what can we predict about the outcome?.
If we want to know the effects of a predictor, we may want to control for covariates with a multivariate model, in order to estimate “direct” effects; leave covariates out, in order to estimate “total” effects (direct plus indirect); or take an intermediate strategy.
These address different versions of the question: if we change the focal parameter (possibly allowing it also to affect other covariates), what is the expected effect on the outcome variable.

\jd{Is there a better place to cite this?}
Shi et al. \citep{shi_evidence_2017}, for example, used multivariate effects plots to visualize estimated direct effects of different predictors in a paper that compared the difference in sexual risk behaviors between circumcised and uncircumcised men.

Producing this sort of outcome plot has some challenges, including:
\begin{enumerate}
\item choosing the \emph{reference point} for non-focal predictors in multivariate models
\item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals in effects plots
\item biases induced by non-linear transformations of the response variable in generalized linear models (especially generalized mixed models).
\end{enumerate}

Representative values for a focal predictor are generally chosen using quantiles, or equally spaced values, for a continuous predictor; or an exhaustive set of levels for a categorical predictor.
The central estimates are then calculated by holding the non-focal predictors at a reference point (a value chosen for a non-focal predictor) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}. These values have been called: \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package} In this article, we refer to these quantities as the \emph{central estimates} of the outcome.

In a model with non-focal predictors such as multivariate models, reference points can be chosen as the average of the non-focal \emph{linear predictor variables} -- we call this approach \emph{mean-based} reference point and is currently not implemented in commonly used \proglang{R} software packages. We introduce an alternative choice for the reference point.

For a linear model, the averaging is done on the linear scale, i.e., linear averaging. As a result, the model-center estimates (made using the mean-based approach) are unbiased. 
However, in a model with non-linear link function, this is not usually true. When averaging is done on a separate link scale, the mean of the estimates is not the same as the estimate at the mean point. 
This leads to bias: in this case a systematic difference between the values seen on average for a given value of the focal predictor and the value predicted by the mean-based approach. 
An alternative to the mean-based reference point is the \emph{observed-value-based} approach, discussed later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.

This article will discuss and implement various approaches for computing predictions and effects plots. We further explore and demonstrate, using simulated data, approaches for correcting bias in central estimates for generalized models involving non-linear link functions, including models with random effects. The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects plots (in our sense), and 2) a robust way to correct for non-linear averaging bias in generalized (mixed) models.

@@

In order to discuss the statistical background and mathematical formulation of the proposed approaches, we need to understand and formally define a number of terms, some of which have been introduced in the previous section:

@@

\item [Model matrix] Refers to the design matrix whose rows include all combination of input variables. Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$. Suppose we want to model the household size (\code{hhsize}) as function of these household characteristics, i.e., $$\mathrm{hhsize} = \beta_0 + \beta_1\times\mathrm{income} + \beta_{2[r]}\times\mathrm{religion} + \mathrm{error}.$$ The model matrix corresponding to this model is given by
$$\begin{bmatrix}{}
 Intercept & income & religionJew & religionMuslim \\
 1 & 50 & 0 & 0 \\
  1 & 100 & 0 & 1 \\
  1 & 77 & 1 & 0 \\
\end{bmatrix}.$$ The first column represents the constant term in our model, $\beta_0$. For continuous input variables, the representation in the model matrix is the same as the corresponding input variables (for example second column representing income). For categorical variables, however, by default, the model matrix  creates additional dummy variables using the reference cell parameterization. This means that, if an input variable has $L$ factor levels, then there will be $L-1$ dummy columns representing all but the first level created in the model matrix. In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.

@@

\item [Reference point] The values (or sets of values) chosen for non-focal predictors, when estimating the predictions and effects. Typically the center point, but can also be chosen as a baseline value or as a mean across categories. We will also discuss using a set of quantiles or observations as a reference.

\item [Anchor] The value chosen for the focal predictor when estimating effect-style confidence intervals. The anchor choice does not affect the central estimates, nor prediction-style. Often chosen as the center point of the linear predictor variables corresponding to the focal predictor.

\item [Prediction-style and effect-style plots] A prediction-style plot is about predicting observations for a given value of the focal predictor. For this, we want to use the classic curved confidence intervals. An effect-style plot attempts to visualize the effect of a focal predictor and are characterized with narrower confidence intervals. Unlike a prediction-style plot, where the confidence intervals capture the uncertainty associated all predictors in the model, the effect-style plot focuses on uncertainty associated with the focal predictor only and depend on the anchor.

>>>>

Plots of predicted values of an outcome against predictors (often called effect plots or prediction plots, generalized here as “outcome plots”) are often a useful way to summarize the results of a statistical model. These can be used to the shape of model response and associated uncertanties, and may be easier to interpret than tables of coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.

To make an outcome plot, we use a \emph{focal} predictor on the x-axis and values of the outcome variable on the y-axis. The resulting plot will depend on choices about other (non-focal) predictors. In an ordinary linear model, the non-focal choices may have a simple additive effect on the central estimate. However, when the focal predictor has interactions or non-linear response functions (e.g., spline or polynomial), non-focal choices can also affect the slope of the estimate. Additional challenges arise when dealing with “mixed” models, which incorporate random effects.

@@

The other distinction relates not to the method of calculating CIs, but to the model chosen for the plot. 
In general, if we want to predict based on a focal parameter, we are likely to want to fit a univariate model that only contains terms related to that predictor, thus addressing the question: if we know (only) the focal parameter, what can we predict about the outcome?
If we want to know the effects of a predictor, we may want to control for covariates with a multivariate model, in order to estimate “direct” effects; leave covariates out, in order to estimate “total” effects (direct plus indirect); or take an intermediate strategy.
These address different versions of the question: if we change the focal parameter (possibly allowing it also to affect other covariates), what is the expected effect on the outcome variable (see, e.g., \citep{shi_evidence_2017}).

We discuss the idea of a model “center” and show how it simplifies thinking about outcome plot choices, particularly in the case of ordinary linear models. We also discuss biases that arise from non-linear link functions in generalized linear models, and how such biases can be addressed.

@@

\jd{We should make this a box.}

@@

@@

\item [Reference point] The values (or sets of values) chosen for non-focal predictors, when estimating the predictions and effects. Often based on the model center, can also use chosen baseline values, or an unweighted mean across categories for a categorical variable. We will also discuss using a set of quantiles or observations as a reference.

\item [Anchor] The reference value chosen for variables related the focal predictor when estimating effect-style confidence intervals. The anchor choice does not affect the central estimates, nor prediction-style confidence intervals. Often based on the model center, can also be chosen.
